Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{{DISTRIBUTION STATEMENT A. Approved for public release.} }

### 2021-05-14 paper/synthesis.tex
Worsmithing the Synthesis section

Synthesis maps from model and specifications to code. The synthesis algorithm traverses the system architecture looking
for occurrences of filter and monitor specifications. For each such 
occurrence it generates a CakeML program. In the following, we examine both filter and monitor synthesis. The latter is typically much more involved, and we will devote more attention to it.

@@

A filter is intended to be very simple; it is expected to have one
input port and one output; messages on the input that the filter
policy admits pass unchanged to the output port; all others are not
passed on. We have explored in our work two kinds of filter. In the
first, a relatively shallow scan of the input buffer can enforce the
policy. For example, we have used the expressive power of Contiguity
Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints
on GPS coordinates in UxAS messages. On the other hand, the second
kind of filter will parse the input buffer into a data structure
specified in AGREE and apply an user-defined \emph{well-formedness}
property, also specified in AGREE, to the data. This allows
arbitrarily complex well-formedness checking.

The decision of a filter is made and performed within one thread
invocation. Thus, in its given time slice, the filter does the following:

@@

The role of partiality should be emphasized: steps 2 and 3 can fail;
the data might not be parseable or the wellformedness computation
could be badly written and fail at run time. In such cases, the filter
should recover and yield control without passing the input onwards. In
these cases, the filter is behaving as it should, but there are also
cases in which a correctly specified filter would not meet its
specification at runtime. This situation arises when the
filter \emph{ought} to accept a message, but lack of resources result
in the filter failing to do so. Examples of this would be, for
example, if the parse of a message needed more space than allocated;
another example would be if the timeslice provided by the scheduler is
too small for the wellformedness computation to finish.

@@

Monitors, since they are intended to track and analyze the externally visible behavior of system components through time, require more computational features than filters. In particular, our conception of a monitor is a predicate on its input (and output) streams, being able to access the value of a stream at any earlier point in time,  if necessary. Thus monitors commonly use state to keep track of earlier values, unlike filters which are stateless. This reasoning leads us to specify the  computation for a monitor component  by a step function of the following form:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
Thus, in its given timeslice, a monitor evaluates the \konst{stepFn} on its inputs and the current values of the state variables. In detail, it takes the following steps:

@@

The scheduler \emph{activates} components in some order.It is an obligation on the system that the scheduler follows some sensible partial order of component activation and allows each component sufficient time for its computation.
 Activating a monitor component takes the form of the following pseudo-code: 
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stepFn} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\paragraph{Initialization}
A monitor may need to accumulate a certain minimum number of observations before being able to make a meaningful assessment of behavior. Until that threshold is attained, the monitor is essentially in a kind of \emph{initialization} phase. In order for correct code to be generated, monitor specifications need to spell out the values of output ports when in such initialization phases. For example, suppose a monitor does some kind of differential assessment of inputs at adjacent timeslices, alerting when (say) the measured location of a UAV at times $t$ and $t+1$ is such that the distance between the two locations is unusually large. Such a monitor needs two measurements before making its first judgement, but at the time of its first output, only one measurement will have been made. The specification has to explicitly say what the correct first output is.

>>>>

Synthesis maps from model and specifications to code. Our synthesis algorithm traverses the system architecture looking
for occurrences of filter and monitor specifications. For each such 
occurrence, the synthesis algorithm generates a CakeML program. In the
following, we examine both filter and monitor synthesis. The latter is
typically much more involved, and we will thus devote more attention to it.

@@

A filter is intended to be very simple,  It is expected to have one
input port and one output; messages on the input that the filter
policy admits pass unchanged to the output port; all others are not
passed on. We have explored two kinds of filter in our work to date. In the
first, a relatively shallow scan of the input buffer can enforce the
desired policy. For example, we have used the expressive power of Contiguity
Types \cite{contiguity-types} to enforce \emph{lightweight} bounds constraints
on GPS coordinates in UxAS messages.  The second
kind of filter parses the input buffer into a data structure
specified in AGREE and applies a user-defined \emph{well-formedness}
property, also specified in AGREE, to the data. This allows
arbitrarily complex wellformedness checking.

A filter performs its processing and makes its decision all within one thread
invocation. Thus, in its given time slice, the filter does the following:

@@

The role of partiality should be emphasized.  Steps 2 and 3 can fail;
the data might not be parseable or the wellformedness computation
could fail at run time. In such cases, the filter
should recover and yield control without passing the input to the output. In
these cases, the filter is behaving as it should, but there are also
cases in which a correctly specified filter would not meet its
specification at runtime. This situation arises when the
filter \emph{ought} to accept a message, but lack of resources result
in the filter failing to do so. Examples of this would be, for
example, if the parse of a message needed more space than allocated;
another example would be if the timeslice provided by the scheduler is
too short for the wellformedness computation to complete.

@@

Monitors require more computational features than filters, as they are
intended to track and analyze the externally visible behavior of
system components through time.  Our conception of
a monitor is a predicate on its input (and output) streams, where the
monitor is able to access the value of a stream at any earlier point in time,  if necessary. Thus monitors commonly use state to keep track of earlier values, unlike filters which are stateless. This reasoning leads us to specify the  computation for a monitor component  by a step function of the following form:
\[
\konst{stepFn} : \mathit{input} \times \mathit{stateVars} \to \mathit{stateVars} \times \mathit{output}
\]
Thus, in its given timeslice, a monitor evaluates the \konst{stepFn} on its inputs and the current values of the state variables. In detail, it takes the following steps:

@@

The scheduler \emph{activates} components in some order.  It is an obligation on the system that the scheduler follows some sensible partial order of component activation and allows each component sufficient time for its computation.
 Activating a monitor component takes the form of the following pseudo-code: 
\[
\begin{array}{ll}
 \mathit{(i_1,\ldots)} & = \konst{readInputs}(); \\
 (v_1,\ldots) & = \konst{readState}() ; \\
 ({v_1}',\ldots), ({o_1}',\ldots) & = \konst{stepFn} ((i_1,\ldots),(v_1,\ldots)) ; \\
 \multicolumn{2}{l}{\konst{writeState}({v_1}',\ldots);} \\
 \multicolumn{2}{l}{\konst{writeOutputs}({o_1}'\ldots);} \\
\end{array}
\]

\paragraph{Initialization}
A monitor may need to accumulate a certain minimum number of
observations before being able to make a meaningful assessment of
behavior. Until that threshold is attained, the monitor is essentially
in a kind of \emph{initialization} phase. In order for correct code to
be generated, monitor specifications need to spell out the values of
output ports when in such initialization phases. For example, suppose
a monitor does some sort of differential assessment of inputs at
adjacent timeslices, alerting when (say) the measured location of a
UAV at times $t$ and $t+1$ is such that the distance between the two
locations is unusually large. Such a monitor needs two measurements
before making its first judgement, but at the time of its first
output, only one measurement will have been made. In this case, the
specification has to explicitly say what the correct first output is.
